Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-03-07 numerics_paper/version2/numerical_chj.tex
changes requested by Daviid

%\newdefinition{rmk}{Remark}
\newtheorem{remark}{Remark}

@@

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solvers are used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe
based on a local measure of the entropy residual, 
as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.

@@

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but 
(due to the use of excessive numerical dissipation)
is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Recall that $Q=[h,hu,hv]^t$. We use the total energy as entropy; that is, we consider 
\begin{align*}
  \eta(\bfq)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(\bfq)=
    \frac{\eta(\bfq)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\begin{remark}[About the design philosophy.]
  The work by Tadmor \cite{tadmor2003entropy} is based on imposing condition \eqref{ent_ineq} as equality. 
  By doing this, the method is said to be entropy conservative. To improve the nonlinear 
  stability properties of the numerical scheme, a consistent (and potentially high-order) 
  artifical entropy dissipative operator is included. 
  In this work, we add dissipation on the conserved variables based on 
  a measurement of the entropy residual 
  (as in \cite{guermond2011entropy,guermond2018second,guermond2018well}).
  Doing this, however, does not guarantee \eqref{ent_ineq}. 
  Indeed, some linear stabilization techniques that add artificial dissipation on the conserved 
  variables are known to produce entropy; see for example \cite{ern2013weighting,kuzmin2020entropy}. 
  To guarantee \eqref{ent_ineq}, we add extra dissipation on the conserved variables via 
  \eqref{lambda_min}. 
  Doing this counteracts entropy production created by any component of the Riemann solver;
  see for example \cite{kuzmin2020entropy} (in the context of $C^0$ finite elements). 
\end{remark}

>>>>

%\newdefinition{rmk}{Remark}
\newtheorem{remark}{Remark}
\newtheorem*{remark*}{Remark}{}

@@

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe
based on a local measure of the entropy residual, 
as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.

@@

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but 
(unless the mesh is highly refined)
is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

@@

\subsection{The entropy residual}
Let $\entvar(q)$ and $\entflux(q)$ denote the entropy variable and entropy flux.
Recall that $q=[h,hu,hv]^T$. We use the total energy as entropy; that is, 
\begin{align*}
  \eta(q)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(q)=
    \frac{\eta(q)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(q) \cdot \left[ \frac{\partial q}{\partial t} + \nabla\cdot \bff(q)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(q)}{\partial t} + \entvar(q) \cdot \nabla\cdot \bff(q)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(q)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(q) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(q) \cdot \nabla\cdot \bff(q) - \nabla\cdot\entflux(q) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $q$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(q)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(q) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(q)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(q)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\begin{remark*}[About the design philosophy.]
  The work by Tadmor \cite{tadmor2003entropy} is based on imposing condition \eqref{ent_ineq} as equality. 
  By doing this, the method is said to be entropy conservative. To improve the nonlinear 
  stability properties of the numerical scheme, a consistent (and potentially high-order) 
  artifical entropy dissipative operator is included. 
  In this work, we add dissipation on the conserved variables based on 
  a measurement of the entropy residual 
  (as in \cite{guermond2011entropy,guermond2018second,guermond2018well}).
  Doing this, however, does not guarantee \eqref{ent_ineq}. 
  Indeed, some linear stabilization techniques that add artificial dissipation on the conserved 
  variables are known to produce entropy; see for example \cite{ern2013weighting,kuzmin2020entropy}. 
  To guarantee \eqref{ent_ineq}, we add extra dissipation on the conserved variables via 
  \eqref{lambda_min}. 
  Doing this counteracts entropy production created by any component of the Riemann solver;
  see for example \cite{kuzmin2020entropy} (in the context of $C^0$ finite elements). 
\end{remark*}
